Composite Bayesian inference

### 2019-04-08 papers/composite_inference/draft.8apr19.tex
Modifs apr 8

%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{proposition}[theorem]{Proposition}

@@

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}
%\newcommand{\matphi}{\boldsymbol{\Phi}} 
%\def\p{{\bar{\mathbf{p}}}}
%\def\q{{\bar{\mathbf{q}}}}

@@

\begin{abstract}
We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents. The resulting class of predictive models, which we call ``composite Bayesian'', is a middle way between purely generative and purely discriminative models. This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning: in the latter case, they may be tuned so as to maximize training likelihood, yielding an easy-to-solve convex problem. We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
\end{abstract}

@@

Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a target variable conditional on the data, in contrast with a {\em generative} model, which describes the conditional distribution of the data given the target variable.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).

Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves. Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth. A discriminative model can learn a threshold that correctly classifies most pans in practice, but is unable to see that, say, a blender does not match either category. The impossibility of such confirmatory analysis strongly limits interpretation if outlier inputs are to be expected.

There is growing awareness that AI should be interpretable in life-impacting applications~\cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision making processes. For instance, in medical diagnosis, automated disease predictions should be corroborated by individualized findings to be communicable to clinicians. In other words, there is no interest for ``black boxes'' but for AI systems that can justify their opinions just like human experts. On the other hand, clinicians have long used reference range analysis for diagnosis, which stems from simple univariate generative models.

@@

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.25\textwidth]{generative.pdf}\label{fig:generative}}
\hspace*{.2\textwidth}
\subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
\caption{Directed graphs representing generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the target variable, the data, and the model parameters. Note the marginal independence of data and parameters in the discriminative model.}
\label{fig:graph_comparison}
\end{center}
\end{figure}

This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both computationally efficient and interpretable, resulting in a class of hybrid models. The key idea is to combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data. Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the target variable. The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior. This may be understood as a probabilistic version of boosting, or as a {\em divide and conquer} approximation to the intractable Bayesian posterior.

%Importantly, predictions made in such a way are interpretable owing to their underlying generative nature.

When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL). As discussed in \cite{Varin-11} and the references therein, CL was developed as a surrogate of traditional likelihood for parameter estimation. While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency.

With the opinion pooling interpretation in mind, the weights assigned to the different features may be optimized for prediction performance in a typical discriminative learning scenario. As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions. The likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically involves a double training scheme: a generative training stage (to learn the feature-based likelihood parameters) followed by a discriminative training stage (to learn the feature weights in aggregation).

Technical details are given in the remainder. We warn the reader that slightly abusive mathematical notations are used deliberately for the sake of clarity in places where we think that they are unambiguous.

@@

If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. One that is appealing as it tends to produce single peaked distributions is the (generalized) log-linear opinion pool:
\begin{equation}
\label{eq:log_pool}
p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n f(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is a prior distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights. Note that it is possible to have the weights depend on the data~$\y$, but this would make the method considerably more complex and less interpretable. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.

%This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are upper bounded. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|\y)\leq p_\lambda(x_2|\y)$. Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and less interpretable.

%Strictly positive weights guarantee the so-called 0/1 forcing property, that is, if an hypothesis~$x$ has zero likelihood according to at least one agent, then its consensus probability vanishes too.

% Not clear yet as to what a negative weight could mean!

@@

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule as it yields the form: 
$$
p_\blambda(x|\y)\propto \pi(x) L^c_\blambda(x),
$$
where the distribution $\pi(x)$ plays the role of a prior, and the function:
\begin{equation}
\label{eq:comp_lik}
L^c_\blambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays that of a likelihood function. The expression~(\ref{eq:comp_lik}) happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}. The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}. The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features.

CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$. While the CL derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method, however unitary CL weights do not guarantee optimal prediction performance.

@@

If training data is available, the generative parameters may be set in a first stage using standard estimation techniques. Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to composite likelihood one of the same techniques as for traditional likelihood \cite{Berger-99}. We focus in the sequel on tuning the composite weights.

@@

In the absence of knowledge, uniform weights should be chosen under a rule of indifference. CL is then a scaled version of na\"ive Bayes likelihood, the unique weight value being irrelevant to the maximum CL estimator (MCLE). To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, as proposed in previous attempts at Bayesian inference from composite likelihood.

% in attempts to match frequentist and Bayesian notions of uncertainty.

Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one. This is motivated by the opinion pooling perspective and a characterization theorem \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights. The external Bayesian  property essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently. The log-linear pool with unit sum weights also happens to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.

@@

Unit sum weights, however, implicitly assume maximum redundancy between features. Consequently, they tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}), meaning reduced prediction confidence when features are weakly correlated, as is desirable in general. An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}. This suggests a learning approach if possible, which, as we shall see, is unlikely to pick uniform or unit sum weights.

@@

Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
\begin{equation}
\label{eq:train_likelihood}
\max_{\blambda\succeq 0} U(\blambda),
\qquad \text{with} \quad
U(\blambda) \equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
\end{equation}

From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\blambda(x|\y)$ spanned by the weights~$\blambda$ is the exponential family with natural parameter~$\blambda$ and basis functions given by the feature-based log-likelihoods:
$$
p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - a(\blambda,y)],
$$
with:
$$
\ell_i(x,\y) = \log f(z_i|x),
\qquad
a(\blambda, \y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
$$

It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence the learning can be implemented using a standard convex optimization scheme such as the limited-memory BFGS algorithm \cite{Byrd-95}. See Appendix~\ref{app:training} for some implementation details.

Because some positive weight constraints may turn out inactive, learning has a natural tendency to produce sparse weights. As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\textwidth]{disc_weight_plot.pdf}
  \end{center}
\caption{Distribution of optimal composite weights vs individual feature discrimination power in a binary classification task (breast cancer UCI dataset). Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.}
\label{fig:disc_weight_plot}
\end{figure}

As is customary and generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$, enforcing balanced classes if the prior is uniform. Also, some regularization may be needed for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.

@@

%D(p\|p_0) = \sum_{x\in{\cal X}}\sum_{\y\in{\cal Y}} p(x,\y) \log \frac{p(x,\y)}{p_0(x,\y)},
Exponential family properties also imply that learning via likelihood maximization~(\ref{eq:train_likelihood}) is dual to an $I$-projection \cite{Csiszar-84}: 
\begin{equation}
\label{eq:maxent}
\min_{p\in{\cal P}} D(p\|p_0),
\qquad \text{with} \quad
p_0(x,\y) \equiv \pi(x)h(\y),
\end{equation}
subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where $h(x,\y)$ is the joint empirical distribution of targets and data examples, $h(\y)$ is the corresponding marginal data distribution, and ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$ with ${\cal Y}\equiv\{\y_1,\ldots,\y_N\}$. Note that the constraints implicitly approximate the data marginal and the mean log-likelihood values by their empirical estimates, confirming the need for regularization in practice.

The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized, hence the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}. The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature. This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a truthworthy upper bound.

@@

An insightful analysis given in \cite{Grunwald-04} shows that the $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss. Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is inexact due to the limited training set size. Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by the underlying knowledge, which in this case may be interpreted in terms of maximum feature description length.

@@

So far, we have assumed that each agent is given a single weight in the opinion aggregation~(\ref{eq:log_pool}). Let us now consider a more general pooling operator where agents can be weighted depending on the target variable:
\begin{equation}
\label{eq:super_pool}
p_\lambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
\end{equation}
where $x_0$ is some fixed target value, which we call the reference. As discussed in Appendix~\ref{app:super}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting functions~$w_i(x)$ sum up to one uniformly,
$$
\forall x\in{\cal X},\quad
\sum_{i=1}^n w_i(x) = 1,
$$
hence providing a sensible default rule to tune the weights. Note that (\ref{eq:super_pool}) amounts to a log-linear pool if the weights are target-independent as the reference is then effectless. Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
\begin{equation}
\label{eq:super_comp_lik}
L^{sc}_\blambda(x) \equiv 
\prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)}
\end{equation} 
plays the role of a standard likelihood at prediction time. We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs. SCL further generalizes CL since $L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.

Conversely, when each target assigns a weight one to a single feature and zero to all the others, SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}. While our derivation stems from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.

%Case of a single nonzero agent per feature. Then SCL is equivalent to the class-specific likelihood derived in \cite{Baggenstoss-03} based on a generative model selection argument  Our derivation is much different but interesting to see that we end up with a more general form.

A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some target values~$x$. It is possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs. Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence on the predictive distribution.

>>>>

@@

\def\x{{\mathbf{x}}}
\def\y{{\mathbf{y}}}
\newcommand{\blambda}{{\boldsymbol{\lambda}}}
\newcommand{\Blambda}{{\boldsymbol{\Lambda}}}
\newcommand{\bell}{{\boldsymbol{\ell}}}
\newcommand{\E}{\mathbb{E}}

@@

\begin{abstract}
We revisit and generalize the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents. The resulting class of predictive models, which we call ``composite Bayesian'', is a middle way between purely generative and purely discriminative models. This perspective gives insight to choose the weights associated with composite likelihood, either {\em a priori} or via learning; in the latter case, they may be tuned so as to minimize training cross-entropy, yielding an easy-to-solve convex problem. We argue that composite Bayesian inference trades off between interpretability and prediction performance, both of which are crucial to many artificial intelligence tasks.
\end{abstract}

@@

Textbook statistical inference (frequentist or Bayesian) rests upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this double requirement may be very challenging for multidimensional data, other inference models have been developed in applied science: deliberately misspecified generative models, as in quasi-likelihood \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01} methods; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the conditional distribution of the target variable given the data, while a {\em generative} model describes the joint distribution of the target and the data.}, which currently dominate the field of artificial intelligence (AI) and typically require supervised learning on large datasets -- these include many classical (shallow) learning \cite{Ho-95,BergerA-96,Vapnik-00,Rasmussen-06} and deep learning \cite{Lecun-15,Goodfellow-16} techniques (with the exception of deep belief networks \cite{Hinton-06}).

In a determined universe of possibilities, discriminative models can map data to predictions as well as intelligent beings or even better, however they lack introspection in the sense that they cannot {\em test} their predictions. Consider as a straightforward example deciding whether an object is a sauce pan or a frying pan based on depth. A two-class discriminative model can learn a threshold that correctly classifies most pans in practice, but will confidently classify a blender as a sauce pan, while linear discriminant analysis\footnote{Despite the name, linear discriminant analysis is based on a generative model.}, for instance, could hint that a blender is a poor match to both pan categories. Inability to ``confirm'' or ``justify'' their predictions makes discriminative models difficult to interpret.

%Discriminative models, however, lack the notion of {\em likelihood}: they cannot assess whether the data is consistent with their own prediction, however accurate, and in this sense do not have the ability to ``justify'' themselves. Consider as a straightforward example the problem of deciding whether an object is a sauce pan or a frying pan based on depth. A discriminative model can learn a threshold that correctly classifies most pans in practice, but is unable to see that, say, a blender does not match either category. The impossibility of such confirmatory analysis strongly limits interpretation if outlier inputs are to be expected.

There is growing awareness that AI should be interpretable in life-impacting applications \cite{Molnar-18}, where it is (and perhaps should remain) confined to a supportive role in human-driven decision-making processes, if only because decisions need to be explained and justified to other humans, while determining the set of possible outcomes may also be part of the process. For instance, in medical diagnosis, automated disease predictions need to be corroborated by individualized findings to be taken into account. In other words, there is little clinical value in ``black boxes'' and certainly more in automated methods that can justify their outputs just like human experts. One such method that has long been used by clinicians is the familiar reference range analysis, which stems from simple generative models.

@@

\begin{figure}[!ht]
\begin{center}
\subfigure[Generative model]{\includegraphics[width=.25\textwidth]{generative.pdf}\label{fig:generative}}
\hspace*{.2\textwidth}
\subfigure[Discriminative model]{\includegraphics[width=.25\textwidth]{discriminative.pdf}\label{fig:discriminative}}
\caption{Directed graphs representing a generative and discriminative models, where $X$, $Y$ and $\theta$ respectively denote the target variable, the data, and the model parameters. Note the marginal independence of data and parameters in the discriminative model.}
\label{fig:graph_comparison}
\end{center}
\end{figure}

This note advocates probabilistic opinion pooling \cite{Genest-86} as a natural way to merge discriminative and generative modeling approaches in order to make predictions that are both accurate and interpretable. The key idea is to combine multiple low-dimensional generative models corresponding to different pieces of information extracted from the input data. Each such model acts as an isolated ``agent'' that uses a single feature to express a testable opinion in the form of a likelihood function of the target variable. The agent opinions are then aggregated into a unique predictive probability distribution analogous to a Bayesian posterior. This strategy may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior.

%Importantly, each agent opinion is justifiable as it is based on a generative model.

When choosing the aggregation method as a log-linear pool, the predictive distribution turns out to be proportional to a quantity known in computational statistics as composite likelihood (CL), see \cite{Varin-11} and the references therein. CL was developed as a surrogate of traditional likelihood for parameter estimation. While maximum CL does not inherit the general property of maximum likelihood to achieve asymptotical minimum variance, it is asymptotically consistent under mild conditions \cite{Xu-11} (see Appendix~\ref{app:frequentist} for a basic weak consistency proof), and may offer an excellent trade-off between computational and statistical efficiency in practice.

With the opinion pooling interpretation in mind, the weights assigned to the different features in~CL may be optimized for prediction performance in a typical supervised learning scenario. As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions. The likelihood functions themselves may need to be pre-trained, therefore the composite inference approach typically involves a double training scheme: a generative training stage (to learn the feature-based likelihood parameters) followed by a discriminative training stage (to learn the feature weights in aggregation).

The opinion pooling framework also suggests a generalization of CL, which we call {\em super composite likelihood}, in which the features may be weighted depending on the target values, leading to a more flexible predictive model that can be trained similarly to CL.

%Technical details are given in the remainder.

The reader is warned that slightly abusive mathematical notation is sometimes used in the remainder. This is done deliberately for the sake of clarity in places where we think that notation can be lightened without raising ambiguity.

@@

If we see the likelihoods as Bayesian posterior distributions corresponding to {\em uniform} priors, this is a problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. One that is appealing as it tends to produce single peaked distributions is the log-linear opinion pool\footnote{When $\pi(x)$ is not uniform, it is also called {\em generalized} log-linear pool, however this terminology could be confusing here since we will present another generalization.}:
\begin{equation}
\label{eq:log_pool}
p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n f(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is a reference distribution, and $\blambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights. Note that the weights could depend on the data~$\y$, but this would make the method considerably more complex and less interpretable. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less probable than an outcome $x_2$, this should be reflected by the consensus probabilities, {\em i.e.}, we should have $p_\blambda(x_1|\y)\leq p_\blambda(x_2|\y)$.

@@

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule as it yields the form: 
$$
p_\blambda(x|\y)\propto \pi(x) L^c_\blambda(x),
$$
where the distribution $\pi(x)$ plays the role of a prior, and the function:
\begin{equation}
\label{eq:comp_lik}
L^c_\blambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays that of a likelihood function, and happens to be known in computational statistics as a {\em marginal composite likelihood} \cite{Varin-11}. The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{app:conditional}.

The computational advantage of CL over genuine likelihood stems from the fact that it is more efficient to evaluate the marginal distributions of each feature than the joint distribution of all features. CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, often referred to as {\em na\"ive Bayes} method in the machine learning literature, which corresponds to the special case of unitary weights, $\blambda\equiv 1$. Since log-linear opinion pooling does not assume feature independence, it yields both an alternative interpretation and a generalization of na\"ive Bayes, whereby unitary feature weights may be used by default but do not guarantee optimal prediction performance.

@@

If training data is available, the generative parameters may be set in a first stage using standard estimation techniques. Otherwise, they may be considered as nuisance parameters and eliminated at prediction time by applying to CL one of the same techniques as for traditional likelihood \cite{Berger-99}. We focus in the sequel on tuning the composite weights.

@@

In the absence of knowledge, uniform weights may be chosen under a rule of indifference. CL is then a scaled version of na\"ive Bayes likelihood, the unique weight value being irrelevant to the maximum CL estimator (MCLE). To get meaningful predictive probabilities, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic MCLE variance matrix \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, as proposed in previous attempts at Bayesian inference from composite likelihood.

Alternatively, a simple agnostic recommendation we believe to be useful for composite weights is to sum up to one. This is motivated by a characterization theorem for opinion pooling \cite{Genest-86,Genest-86b}: the only {\em externally Bayesian} pooling operator (for which the consensus probability of an outcome only depends on the agent probabilities of the same outcome) is the log-linear opinion pool with unit sum weights. Externally Bayesian essentially means that the consensus should not change if an agent opinion is taken into account by the other agents as opposed to being stated independently. The log-linear pool with unit sum weights also turns out to be an optimal {\em compromise} in the sense that it minimizes the average inclusive Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}.

@@

Unit sum weights, however, implicitly assume maximum redundancy between features, hence tend to produce conservative predictive probabilities (see Appendix~\ref{app:frequentist}). This means low prediction confidence (large credibility sets) when features are weakly correlated, as is desirable in general. Conversely, unitary weights ($\blambda\equiv 1$) as in Na\"ive Bayes assume independent features and may thus yield over-confident predictions. An ideal method to weight features is one that can capture feature redundancy and balance it with feature relevance \cite{Peng-05}. This suggests a learning approach whenever feasible.

%, which, as we shall see, is unlikely to pick uniform or unit sum weights.

@@

Assuming a training dataset ${\cal D}=\{(x_k,\y_k), k=1,\ldots,N\}$, possibly the same as for generative pre-training, the most direct way to learn the composite weights is to maximize their likelihood under the composite predictive distribution (or, equivalently, minimize their ``cross-entropy''):
\begin{equation}
\label{eq:train_likelihood}
\max_{\blambda\succeq 0} U(\blambda),
\qquad \text{with} \quad
U(\blambda) \equiv\sum_{k=1}^N \log p_\blambda(x_k|\y_k).
\end{equation}

From~(\ref{eq:log_pool}), we see that the set of conditional distributions $p_\blambda(x|\y)$ spanned by the weights~$\blambda$ is the exponential family with natural parameter~$\blambda$ and basis functions given by the feature-based log-likelihoods:
$$
p_\blambda(x|\y) = \pi(x) \exp[\blambda^\top \bell(x,\y) - a(\blambda,y)],
$$
with:
$$
\ell_i(x,\y) \equiv \log f(z_i|x),
\qquad
a(\blambda, \y) \equiv \log \sum_{x\in{\cal X}} \pi(x) e^{\blambda^\top \bell(x,\y)}.
$$

It follows that the utility function $U(\blambda)$ in~(\ref{eq:train_likelihood}) is concave as a general property of likelihood in exponential families, hence this learning can be implemented using a standard convex optimization algorithm such as limited-memory BFGS \cite{Byrd-95}. See Appendix~\ref{app:training} for some implementation details.

Because some positive weight constraints may turn out inactive, optimization has a natural tendency to produce sparse weights. As illustrated in Figure~\ref{fig:disc_weight_plot} on the breast cancer UCI dataset\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}, there is no clear relation between optimal weights and feature-level discrimination power, showing that maximum likelihood learning differs radically from univariate feature selection as it implicitly takes feature redundancy into account via joint weight optimization.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\textwidth]{disc_weight_plot.pdf}
  \end{center}
\caption{Distribution of optimal composite weights vs.~individual feature discrimination power in a binary classification task (breast cancer UCI dataset). Discrimination is measured by the maximum KL divergence between class-conditional generative distributions.}
\label{fig:disc_weight_plot}
\end{figure}

As is customary and generally good practice, the training examples may be weighted in~(\ref{eq:train_likelihood}) so that the empirical distribution of classes matches the prior~$\pi(x)$, thus enforcing balanced classes if the prior is uniform. Also, some regularization may be needed for stability: we shall in practice maximize $U(\lambda)-\alpha \|\blambda\|^2$, where $\alpha>0$ is a small damping factor.

@@

Exponential family properties also imply that learning via likelihood maximization~(\ref{eq:train_likelihood}) is dual to an $I$-projection \cite{Csiszar-84}: 
\begin{equation}
\label{eq:i_proj}
\min_{p\in{\cal P}} D(p\|p_0),
\qquad \text{with} \quad
p_0(x,\y) \equiv \pi(x)h(\y),
\end{equation}
subject to the constraints $p(\y)=h(\y)$ and $\E_p[\bell] \succeq \E_h[\bell]$, where $\E_h$ denotes the expectation with respect to the joint {\em empirical} distribution of targets and inputs, $h(\y)$ is the corresponding empirical marginal distribution of inputs, and ${\cal P}$ is the set of joint probability distributions on ${\cal X}\times {\cal Y}$ with ${\cal Y}\equiv\{\y_1,\ldots,\y_N\}$. Note that the problem statement implicitly approximates the true (unknown) joint distribution of $(x,\y)$ by its empirical estimate, hence the need for regularization in practice.

The marginal constraint $p(\y)=h(\y)$ implies that only the conditional distribution $p(x|\y)$ is optimized and that the $I$-projection is equivalent to maximum conditional entropy \cite{BergerA-96}. The mean log-likelihood constraints $\E_p[\bell] \succeq \E_h[\bell]$ encapsulate dependences between the target and the data, and consider a model admissible if it guarantees the same average log-likelihood levels as observed separately for each feature. This is weaker a constraint than assuming that the feature generative models $f(z_i|x)$ are true: it only assumes that the {\em description length} \cite{Grunwald-07} achieved by each feature model is a truthworthy upper bound.

@@

An insightful analysis given in \cite{Grunwald-04} shows that $I$-projection is, under broad conditions, a minimax strategy for prediction according to the log loss. Hence, any $I$-projection is in some sense a best possible predictive model given some arbitrary knowledge exctracted from the training data, which is represented by mean-value constraints and is inexact due to the limited training set size. Optimized composite likelihood only differs from other maximum entropy classifiers such as multinomial logistic regression by its underlying inexact knowledge, which in this case may be interpreted in terms of  feature description length.

@@

So far, we have assumed that each agent is given a single weight in the opinion aggregation~(\ref{eq:log_pool}). Let us now consider a more general pooling operator where agents can be weighted depending on the target variable:
\begin{equation}
\label{eq:super_pool}
p_\blambda(x|\y) \propto \pi(x) \prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)},    
\end{equation}
where~$x_0$ is an arbitrary fixed target value, which we call the reference, and $\blambda: {\cal X}\to \mathbb{R}_+^n$ is a weighting function here. As discussed in Appendix~\ref{app:hetero}, the main reason for introducing the normalization by a reference is to make the pooling externally Bayesian if the weighting function sums up to one uniformly,
$$
\forall x\in{\cal X},\quad
\sum_{i=1}^n \lambda_i(x) = 1,
$$
hence providing a sensible default rule to tune the weights. Note that (\ref{eq:super_pool}) amounts to a log-linear pool if the weights are target-independent as the reference is then effectless. Using the same analogy with Bayes rule as in Section~\ref{sec:bayes_rule}, we see that the quantity:
\begin{equation}
\label{eq:super_comp_lik}
L^{sc}_\blambda(x) \equiv 
\prod_{i=1}^n \left[\frac{\ell_i(x)}{\ell_i(x_0)}\right]^{\lambda_i(x)}
\end{equation} 
plays the role of a standard likelihood at prediction time. We call this form super composite likelihood (SCL) owing to the ``doubly composite'' aspect of combining distinctive feature-target pairs. SCL further generalizes CL since both forms are obviously proportional for target-independent weights, $\lambda_i(x)\equiv \lambda_i$.

%$L^{sc}_\blambda(x)=L^c_\blambda(x)/L^c_\blambda(x_0)$ for target-independent weights $\lambda_i(x)\equiv \lambda_i$, hence it is in this case proportional to CL and yields the same predictive distribution.

Conversely, if the weighting function maps each target to a unit vector ({\em i.e.}, for each target, there is a single feature with non-zero weight and this weight is one), then SCL turns out to be identical to the {\em class-specific} likelihood surrogate derived in \cite{Baggenstoss-03} from a generative model selection argument -- namely, the ``PDF projection theorem'', which in fact boils down to an $I$-projection \cite{Minka-04}. While our derivation stems from a {\em discriminative} model selection approach, it is interesting to see that it includes the class-specific method as a special case, and thus sheds new light on this method.

A compelling advantage of SCL over CL is that it can deal with missing likelihood values, which happen if the feature generative distributions $f(z_i|x)$ are unknown for some target values~$x$. It is possible to assign zero weights $\lambda_i(x)$ to every feature-target pair for which $\ell_i(x)$ is unknown, yet using nonzero weights for other pairs. Likewise, the features for which the reference likelihood $\ell_i(x_0)$ is unknown receive zero weight for all target values, implying that they have no influence whatsoever on the predictive distribution.
